Trending assessment for nowcasts, measurements, and forecasts

### 2024-03-07 sections/03-trending.tex
Update on Overleaf.

Let $F_{t | \tau} (x)$ denote the predicted \ac{cdf} of forecast time $t$ and issue time $\tau$, where the index is analogous to the point prediction notation of Section~\ref{subsec:notation}.
For trending assessment, we compare the probability of positive change with the occurrence of positive changes. 
The probability $p_t$ of a positive change is given by one minus the \ac{cdf} of the prediction at the true value. \unsure{Andere Formulierung für: one minus?}
\unsure{vielleicht sowas: The probability $p_t$ of a positive change is given by the probability that the predicted value at the forecasting time is greater than the known value at the issure time. }
As for the differences in Section~\ref{subsec:notation}, the computation differs slightly for nowcasts and forecasts.
For nowcasts, the probability is calculated by
\begin{equation*}
    p_t = 1 - F_{t+l | t+l} (y_{t})\quad t = 1, \dots, T-l;
\end{equation*}
while for forecasts, it is given by
\begin{equation*}
    p_t = 1 - F_{t+l | t} (y_{t})\quad t = 1, \dots, T-l.
\end{equation*}
Let $z_t$ denote the indicator that the observed change at time $t$ is positive, that is,
\begin{equation*}
    z_t = \ind{\diffyt > 0} \quad t = 1, \dots, T-l.
\end{equation*}
The probabilistic trending evaluation is then based on the probabilities $\mathbf{p} = (p_t)_{t=1}^{T-l}$ and the observed changes $\mathbf{z} = (z_t)_{t=1}^{T-l}$.
The predictive power of $\mathbf{p}$ for $\mathbf{z}$ can be assessed using probabilistic dichotomous forecast evaluation methods.
Dichotomous forecasts predict a binary outcome, for example, a positive or negative change.
Numerical measures for forecast quality are scoring rules.

The \ac{bs} is a widely used scoring rule for dichotomous probabilistic forecasts~\parencite{Brier1950}.
It is defined as
\begin{equation*}
    BS (\mathbf{p}, \mathbf{z}) = \frac{1}{T-l} \sum_{t=1}^{T-l} (p_t - z_t)^2.
\end{equation*}
The \ac{bs} assesses the calibration and sharpness of the forecast and the observation simultaneously~\parencite{Ranjan2010,Mitchell2011}.
Calibration refers to the statistical consistency of forecasts and observations, while sharpness measures the spread of the forecast distribution.
A smaller spread, and thus, higher sharpness, is preferable, as it indicates greater confidence in the prediction.
The basic paradigm of probabilistic forecasting is to maximize sharpness, subject to calibration~\parencite{Gneiting2014}.
These two conflicting aspects cannot be distinguished in the scoring rule.

@@

If forecasts or nowcasts are given as quantiles, $p_t$ can be determined by interpolations among the quantiles.
Let $q_p$ denote the quantiles for forecast time $t+l$ for even-spaced probabilities $p \in \{1/\pmax, \dots, (p-1) / \pmax\}$ ($\pmax \in \N \setminus \{1, 2\}$) and $y_t$ the true value at time $t$.
The quantiles $q_p$ generally differ for each time step, but we omit an index here for ease of notation.
The probability $\pc_t$ of a \textit{negative} change is between $p^{\star}$ and $p^{\star} + 1/\pmax$ for
\begin{equation*}
    p^{\star} = \max \{p \in \{1/\pmax, \dots, (\pmax-1) / \pmax\}: q_p \leq y_t\} , \quad \text{if}\ q_{1/\pmax} \leq y_t \leq q_{1 - 1/\pmax}.
\end{equation*}
Quantiles do not determine the location within the interval $[p^{\star}, p^{\star} + 1/\pmax]$.
Under the assumption of a uniform distribution within the quantile interval, the probability of a negative change is
\begin{equation*}
    \bar{p}_t = \frac{y_t - q_{p^\star}}{\pmax (q_{p^{\star} + 1} - q_{p^{\star}})} + p^{\star}.
\end{equation*}
The approach does not assign probabilities for $y_t$ smaller than the smallest quantile $q_{1/p}$ or larger than the largest quantile.
As a simple extension, we assume that the probability mass is uniformly distributed on an interval of the same length as the nearest interval specified by the quantiles.
This yields
\begin{equation*}
\bar{p}_t = \begin{cases}
    \max \{\frac{1}{\pmax} - \frac{q_{p^\star} - y_t}{\pmax (q_{p2/\pmax} - q_{1/\pmax})}, 0\} &, \text{if } y_t < q_{1/p}, \\
    \min \{\frac{1}{\pmax} - \frac{y_t - q_{(\pmax-1)/\pmax}}{\pmax (q_{(\pmax-1)/\pmax} - q_{(\pmax-2)/\pmax})}, 1\} &, \text{if } y_t > q_{1 - 1/p}, \\
    \frac{y_t - q_{p^\star}}{\pmax (q_{p^{\star} + 1} - q_{p^{\star}})} + p^{\star} &, \text{otherwise.}
\end{cases}
\end{equation*}
The probability of positive change is $p_t = 1 - \bar{p}_t$.

If the true value is given as a distribution because it is still unknown, the probabilities $p_t$ can be computed by integration.
Let for two nowcasts the distributions be given by \acp{pdf} $f_{t+l|t+l}$ and $f_{t|t+l}$ with \acp{cdf}  $F_{t+l|t+l}$ and $F_{t|t+l}$.
Then, the probability of a negative change can be computed by
\begin{align}
    \bar{p}_t
        &= \int_{\begin{subarray}{l}x_1, x_2 \in \R: \\ x_2 < x_1\end{subarray}} f_{t|t+l} (x_1) f_{t+l|t+l} (x_2)  \ \textrm{d} \: x_2 \ \textrm{d} \: x_1 \nonumber\\
        &= \int_{x_1 \in \R} \int_{-\infty}^{x_1} f_{t|t+l} (x_1) f_{t+l|t+l} (x_2)  \ \textrm{d} \: x_2 \ \textrm{d} \: x_1 \nonumber\\
        &= \int_{x_1 \in \R} f_{t|t+l} (x_1) F_{t+l|t+l} (x_1)  \ \textrm{d} \: x_2 \ \textrm{d} \: x_1 \label{eq:trending-probabilistic-pdf}.
\end{align}
Thereby, the distributions are assumed to be independent.
As a Monte Carlo approximation of Equation~\eqref{eq:trending-probabilistic-pdf}, the probability can also be calculated by sampling from $f_{t+l|t+l}$ and $f_{t|t+l}$ and calculating the fraction of negative changes.
For forecasts, the indexes have to be shifted.
If no \acp{pdf} are available, they can be estimated from the \ac{cdf} or quantiles, or the \ac{cdf} or quantiles can be used to generate samples for the Monte Carlo approximation.

>>>>

Let $F_{t | \tau} (x)$ denote the predicted \ac{cdf} of forecast time $t$ and issue time $\tau$, where the index is analogous to the point prediction notation of Section~\ref{subsec:notation}.
For trending assessment, we compare the probability of positive change with the occurrence of positive changes. 
The probability $p_t$ of a positive change is given by one minus the \ac{cdf} of the prediction at the true value. \unsure{Andere Formulierung für: one minus?}
\unsure{vielleicht sowas: The probability $p_t$ of a positive change is given by the probability that the predicted value at the forecasting time is greater than the known value at the issure time. }
As for the differences in Section~\ref{subsec:notation}, the computation differs slightly for nowcasts and forecasts.
For nowcasts, the probability is calculated by
\begin{equation*}
    p_t = 1 - F_{t+l | t+l} (y_{t})\quad t = 1, \dots, T-l;
\end{equation*}
while for forecasts, it is given by
\begin{equation*}
    p_t = 1 - F_{t+l | t} (y_{t})\quad t = 1, \dots, T-l.
\end{equation*}
Let $z_t$ denote the indicator that the observed change at time $t$ is positive, that is,
\begin{equation*}
    z_t = \ind{\diffyt > 0} \quad t = 1, \dots, T-l.
\end{equation*}
The probabilistic trending evaluation is then based on the probabilities $\mathbf{p} = (p_t)_{t=1}^{T-l}$ and the observed changes $\mathbf{z} = (z_t)_{t=1}^{T-l}$.
The predictive power of $\mathbf{p}$ for $\mathbf{z}$ can be assessed using probabilistic dichotomous forecast evaluation methods.
Dichotomous forecasts predict a binary outcome, for example, a positive or negative change.
Numerical measures for forecast quality are scoring rules.
\unsure{wenn ich hier richtig verstehe, macht man hier quasi aus probabilitsicher Vorhersage zu y eine probabilistische Vorhersage dazu, wie wahrscheinlich die Veränderung positiv ist und dann bewerten. Kann man auch hier auch direkt das dem Bewertungsergbnis schießen, wie gut die Vorhersage auf negative Änderung ist? Wenn ja, vielleicht irgendwie erwähnen oder angeben. }
The \ac{bs} is a widely used scoring rule for dichotomous probabilistic forecasts~\parencite{Brier1950}.
It is defined as
\begin{equation*}
    BS (\mathbf{p}, \mathbf{z}) = \frac{1}{T-l} \sum_{t=1}^{T-l} (p_t - z_t)^2.
\end{equation*}
The \ac{bs} assesses the calibration and sharpness of the forecast and the observation simultaneously~\parencite{Ranjan2010,Mitchell2011}.
Calibration refers to the statistical consistency of forecasts and observations, while sharpness measures the spread of the forecast distribution.
A smaller spread, and thus, higher sharpness, is preferable, as it indicates greater confidence in the prediction.
The basic paradigm of probabilistic forecasting is to maximize sharpness, subject to calibration~\parencite{Gneiting2014}.
These two conflicting aspects cannot be distinguished in the scoring rule.

@@

If forecasts or nowcasts are given as quantiles, $p_t$ can be determined by interpolations among the quantiles.
Let $q_p$ denote the quantiles for forecast time $t+l$ for even-spaced probabilities $p \in \{1/\pmax, \dots, (p-1) / \pmax\}$ ($\pmax \in \N \setminus \{1, 2\}$) and $y_t$ the true value at time $t$. \unsure{ist das richtig?}
The quantiles $q_p$ generally differ for each time step, but we omit an index here for ease of notation.
The probability $\pc_t$ of a \textit{negative} change is between $p^{\star}$ and $p^{\star} + 1/\pmax$ for
\begin{equation*}
    p^{\star} = \max \{p \in \{1/\pmax, \dots, (\pmax-1) / \pmax\}: q_p \leq y_t\} , \quad \text{if}\ q_{1/\pmax} \leq y_t \leq q_{1 - 1/\pmax}.
\end{equation*}
Quantiles do not determine the location within the interval $[p^{\star}, p^{\star} + 1/\pmax]$.
Under the assumption of a uniform distribution within the quantile interval, the probability of a negative change is
\begin{equation*}
    \bar{p}_t = \frac{y_t - q_{p^\star}}{\pmax (q_{p^{\star} + 1} - q_{p^{\star}})} + p^{\star}.
\end{equation*}
The approach does not assign probabilities for $y_t$ smaller than the smallest quantile $q_{1/p}$ or larger than the largest quantile.
As a simple extension, we assume that the probability mass is uniformly distributed on an interval of the same length as the nearest interval specified by the quantiles.
This yields
\begin{equation*}
\bar{p}_t = \begin{cases}
    \max \{\frac{1}{\pmax} - \frac{q_{p^\star} - y_t}{\pmax (q_{p2/\pmax} - q_{1/\pmax})}, 0\} &, \text{if } y_t < q_{1/p}, \\
    \min \{\frac{1}{\pmax} - \frac{y_t - q_{(\pmax-1)/\pmax}}{\pmax (q_{(\pmax-1)/\pmax} - q_{(\pmax-2)/\pmax})}, 1\} &, \text{if } y_t > q_{1 - 1/p}, \\
    \frac{y_t - q_{p^\star}}{\pmax (q_{p^{\star} + 1} - q_{p^{\star}})} + p^{\star} &, \text{otherwise.}
\end{cases}
\end{equation*}
The probability of positive change is $p_t = 1 - \bar{p}_t$.

If the true value is given as a distribution because it is still unknown, the probabilities $p_t$ can be computed by integration.
Let for two nowcasts the distributions be given by \acp{pdf} $f_{t+l|t+l}$ and $f_{t|t+l}$ with \acp{cdf}  $F_{t+l|t+l}$ and $F_{t|t+l}$.
Then, the probability of a negative change can be computed by
\begin{align}
    \bar{p}_t
        &= \int_{\begin{subarray}{l}x_1, x_2 \in \R: \\ x_2 < x_1\end{subarray}} f_{t|t+l} (x_1) f_{t+l|t+l} (x_2)  \ \textrm{d} \: x_2 \ \textrm{d} \: x_1 \nonumber\\
        &= \int_{x_1 \in \R} \int_{-\infty}^{x_1} f_{t|t+l} (x_1) f_{t+l|t+l} (x_2)  \ \textrm{d} \: x_2 \ \textrm{d} \: x_1 \nonumber\\
        &= \int_{x_1 \in \R} f_{t|t+l} (x_1) F_{t+l|t+l} (x_1)  \ \textrm{d} \: x_2 \ \textrm{d} \: x_1 \label{eq:trending-probabilistic-pdf}.
\end{align} \unsure{vielleicht eher d(x,y) wegen einem Integralzeichen?}
Thereby, the distributions are assumed to be independent.
As a Monte Carlo approximation of Equation~\eqref{eq:trending-probabilistic-pdf}, the probability can also be calculated by sampling from $f_{t+l|t+l}$ and $f_{t|t+l}$ and calculating the fraction of negative changes.
For forecasts, the indexes have to be shifted.
If no \acp{pdf} are available, they can be estimated from the \ac{cdf} or quantiles, or the \ac{cdf} or quantiles can be used to generate samples for the Monte Carlo approximation.
